---
title: "Getting to know R: more about data types"
date: '2018-08-31'
tags: ["Beginning R", "Data Structures", "Sampling Data Management and Visualization"]
categories: ["R"]
---

```{r setup, message=FALSE}
# .libPaths("P:/RLibrary")
options(stringsAsFactors = TRUE, warn = 0)

knitr::opts_chunk$set(echo = TRUE)
```

## Data Structures 

### Vectors

- numbers, characters, logical
- mixed types of data
- 1 dimension

create this using `c()` or `vector()` 

```{r vector}

vect <- c(letters[1:26]) #  create a vector
vect

length(vect) # how many objects are in the vector

```

There are a couple of different ways to look at the structure of objects

```{r str}
typeof(vect) # type of data in the data structure

class(vect) # get the class of the data

dim(vect) # dimensions

str(vect) # structure of the data

```

#### Challenge
1. create a vector with values 1, 8, 7, 3, 4, 6, 10, 3
2. find the mean, variance, standard deviation, standard error, and range

```{r chall1}

chall2 <- c(1, 8, 7, 3, 4, 6, 10, 3)

mean(chall2) # find the mean
var(chall2) # find the variance
sd(chall2); sqrt(var(chall2)) # standard deviation
sd(chall2)/sqrt(length(chall2))# standard error
range(chall2) # range
min(chall2) ; range(chall2)[1] # min value
max(chall2) ; range(chall2)[2] # max
```


### Matrix

- matrices have two dimensions, rows by columns
- only have one type of data, either character or numeric

```{r matrix, error = TRUE}

mat <- matrix(NA, 5, 4) # create a matrix of NAs that is 5 rows by 4 columns
mat
dim(mat)

2mat <- matrix(0,2,3) # will not work
!mat <- matrix(0,2,3) # will not work
mata-matb <- matrix(0,2,3) # will not work

matTwo <- matrix(0,2,3) # will  work, camel case
mat.Two <- matrix(0,2,3) # will work, but not good coding

c(1,3,7,4,
  2,6,2,1,
  NA,0,0,2)

matChall <- matrix(c(1,3,7,4,2,6,2,1,NA,0,0,2), nrow = 3, ncol = 4, byrow = TRUE)
matChall

dim(matChall)

```

### Data frames

- mixed types of data across columns, must be the same within
- 2 dimensional, rows by columns

```{r dataF }
datA <- data.frame("first_col" = 1:10,
                   "second_col" =  as.character(letters[1:10]),
                   "third_col" = factor(toupper(letters[1:10])),
                   stringsAsFactors = FALSE)

datA

str(datA)

names(datA) # find the column names
names(datA) <- c("firstCol","secondCol,thirdCol")
datA

names(datA)[1] <- c("#firstCol")
datA[,1] # all rows first column
datA[2,1]

datA$`#firstCol` # notice the tick
datA$second_col

colnames(datA)
rownames(datA)
rownames(datA) <- letters[10:19]
rownames(datA) <- NULL
str(datA)

head(datA)
tail(datA)
``` 


## Indexing


```{r}

datA[,1] # first column all rows
datA[2,] # second row all columns

datA$`#firstCol`[c(1,3,10)]

datA[4, "third_col"]

datA[,-2]

datA[,-c(2,3)]

datA[c(1,1,2,2,3,3),]
datA[rep(1:3,2),]
datA[rep(1:3,each = 2),]

names(datA); colnames(datA)

names(datA) == "second_col"

which(names(datA) == "second_col")

datA[which(names(datA) == "second_col")]
datA[-which(names(datA) == "second_col")] # use minus to drop column
datA[which(names(datA) != "second_col")]

datA[,which(names(datA) == "#firstCol" | names(datA) == "third_col")]

datA[,which(names(datA) %in% c("#firstCol","third_col"))]

names(datA) == c("#firstCol","third_col")
```

### Lists

- flexible data structure
- often used to store data
- used with functional programming, `purrr`

```{r lists}
C <- list()
C

C[[1]]  <- c("a","b","c")
C

C[[2]] <- datA

C[["mod1"]] <- lm(mpg ~ hp, data = mtcars)

```

