---
title: "Spatial Autocorrelation"
date: '2018-04-18'
tags: ["Spatial autocorrelation", "Data Seminar"]
categories: ["R"]
---

## Load packages

```{r setup, include=FALSE}
# .libPaths("P:/RLibrary2")
knitr::opts_chunk$set(echo = TRUE)

# install.packages(c("rgeos","rgdal"))
# install.packages("lctools")
# install.packages("spdep")

library(tidyverse)
library(raster)
library(lctools)
library(spdep)
library(nlme)
library(viridis)
```


## What is autocorrelation (spatial or temporal)?

Autocorrelation is the measure of similarity (correlation) between nearby observations (in both time and space)  

### Temporal 

The most recent event is related to the event before and after it


```{r scattercor}
set.seed(1234)

randDat <- data.frame(x = sample(1:1000, 20))

randDat %>% 
  mutate(x_1 = lead(x)) %>% 
  filter(!is.na(x_1)) -> randDat


ggplot(data = randDat) + 
  geom_point(aes(x = x, y = x_1), size = 2) +
  labs(y = "x + 1") +
  theme_classic()

cor(randDat$x, randDat$x_1)
cor.test(randDat$x, randDat$x_1)
```

```{r tempcor}
 randDat %>% 
  arrange(x) %>% 
  mutate(x_1a = lead(x)) %>% 
  filter(!is.na(x_1a)) -> randDat

ggplot(data = randDat) + 
  geom_point(aes(x = x, y = x_1a), size = 2) +
  labs(y = "x + 1") +
  theme_classic()

cor(randDat$x, randDat$x_1a)
cor.test(randDat$x, randDat$x_1a)

acf(randDat$x_1)
```

### Spatial
Spatial autocorrelation is an extension of temporal autocorrelation. Proximal points in space are more similar than distant points.  

Measures of spatial autocorrelation describe the degree that two observations (whether they are points, areas, or raster cells) are similar to each other.  

Spatial autocorrelation in a variable can be exogenous ( caused by another spatially autocorrelated variable like rainfall) or endogenous (caused by the process at play, like spread of disease) 

```{r spatialcor}
p <- shapefile(system.file("external/lux.shp", package="raster"))

p@data$id <- rownames(p@data)
p.points <- ggplot2::fortify(p, region = "id")
head(p.points)

p.df <- full_join(p.points, p@data,by = "id")

ggplot(data = p.df, aes(x = long, y = lat, group = group, fill = NAME_1)) +
  geom_polygon() +
  geom_path(color = "white") + 
  coord_equal() +
  scale_fill_brewer(palette = "Set1") + 
  theme_minimal() + 
  theme(legend.position = "bottom")


Diekirch <- p[p$NAME_1 == "Diekirch",]
Diekirch$value <- c(10,6,4,11,6)
data.frame(Diekirch)

Diekirch.points <- ggplot2::fortify(Diekirch, region = "id")
Diekirch.df <- full_join(Diekirch.points, Diekirch@data,by = "id")


ggplot(data = Diekirch.df, aes(x = long, y = lat, group = group, fill = value)) +
  geom_polygon() +
  geom_path(color = "white") + 
  coord_equal() +
  scale_fill_viridis() + 
  theme_minimal() + 
  theme(legend.position = "bottom")

```

## Adjacent polygons

```{r adjpoly}

w <- spdep::poly2nb(Diekirch)
summary(w)
```

## Compute Morans I

Index of spatial autocorrelation

```{r moransI}
ww <- spdep::nb2listw(w,style = "B")
ww

spdep::moran(Diekirch$value, ww, n= length(ww$neighbours), S0 = spdep::Szero(ww))

moran.test(Diekirch$value, ww, randomisation = FALSE)

moran.mc(Diekirch$value, ww, nsim =99)

```

```{r spdata}
spdata <- read.table("https://stats.idre.ucla.edu/stat/r/faq/thick.csv", header = TRUE, sep = ",")

ggplot(data = spdata) + 
  geom_point(aes(x = east, y = north, color = thick), size = 2) + 
  theme_classic()

coords <- as.matrix(spdata[,c("east","north")])

thicknb<-knn2nb(knearneigh(coords, k = 5), row.names = spdata$X)

moran.test(spdata$thick, nb2listw(thicknb))


cGram<-ncf::spline.correlog(x = spdata$east,
                           y = spdata$north,
                           z = spdata$thick,
                           resamp = 100,
                           quiet = TRUE)

summary(cGram)
plot(cGram)

mod_1 <- lme(fixed = thick ~ soil,
             random = ~ 1|dummy,
             method = "ML",
             data = spdata)
summary(mod_1)

mod_2 <- lme(fixed = thick ~ soil,
             random = ~ 1|dummy,
             correlation = corGaus(1, form = ~ east + north),
             method = "ML",
             data = spdata)
summary(mod_2)
bbmle::AICtab(mod_1, mod_2)

```



