---
title: "Community composition - ordination"
date: '2018-03-30'
tags: ["Community composition", "NMDS", "PCA", "Data Seminar"]
categories: ["R"]
---

## Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# .libPaths("P:\\RLibrary2")

# install.packages("ggrepel")
# install.packages("vegan")
# install.packages("FactoMineR")
# install.packages("factoextra")
# install.packages("lubridate")

library(ggrepel)
library(vegan)
library(FactoMineR)
library(factoextra)
library(lubridate)
library(tidyverse)
library(scales)
```

## Ordination

Read the data in from google drive

```{r loaddata}
id1 <- "1f72ooJvTN9xiFbgMMxL1Kh2ISu9tsJt4"
veg_diversity<- read_csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id1))

id2 <- "1dWDJ-gFKLZ-NXt_Bd0GhxT6Gre3A3GUU"
precip_data<- read_csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id2))

```

```{r datamanip}

glimpse(veg_diversity)

veg_diversity %>% 
  filter(restSite == "Dahms 95") %>% 
  dplyr::select(yearCollect, speciesName, siteYrpercent) %>% 
  group_by(yearCollect) %>% 
  spread(speciesName, siteYrpercent, fill = 0) %>% 
  as.matrix() -> dahms95

## Remove infrequent species 
decostand(dahms95, method = "pa") %>% 
  as.data.frame() %>% 
  summarise_all(sum) %>% 
  gather() %>% 
  filter(value ==1) -> infreq_spp

dahms95 %>% 
  as.data.frame() %>% 
  select(-one_of(c(infreq_spp$key))) %>% 
  as.matrix() -> dahms95

rownames(dahms95) <- dahms95[,1]
dahms95 <- dahms95[,-1]

```

## Conduct the PCA

We will use the `prcomp()` function in Base R

PCA is sensitive to the scale of the data 

If all the data is not measured on the same scale, using covariance means that the result
will be determined by the variable with the greatest value, as it will have the greatest variance

```{r runpca}
dahms95.pca <- prcomp(dahms95, scale = FALSE)
names(summary(dahms95.pca))
summary(dahms95.pca)

```

```{r scree}
data_scree <- data.frame(comp = 1:16, 
                         var = summary(dahms95.pca)$sdev^2)

ggplot(data = data_scree) +
  geom_line(aes(x = comp, y = var)) +
  geom_point(aes(x = comp, y = var), size = 2) +
  theme_classic()

screeplot(dahms95.pca)

fviz_eig(dahms95.pca)

```


## Plotting the PCA

The points on a biplot represent observations positioned  in ordination space by their principal components such that component 1 is represented on the x-axis and component 2 is on the y-axis

interpretation of the ordinatin is dependent on wheter a distance biplot (intersample relationships; scaling = 1) or a correlation biplot (interspecies correlations; scaling = 2) is used, because observation scores are rescaled based on the scaling choice.  

```{r biplots}
fviz_contrib(dahms95.pca, choice = "var", axes = 1:2, top = 25)

fviz_pca_var(dahms95.pca, col.var = "contrib")

fviz_pca_ind(dahms95.pca, col.ind = "cos2", repel = 2)

fviz_pca_biplot(dahms95.pca, axes = 1:2)

# If you want greater control of your plots

scale = 1 # covariance biplot = 1; form biplot = 0
var.scale = scale # scale factor to apply to variables
obs.scale <- 1 - scale # scale factor to apply to observations (rows)

circle.prob = 0.69

nobs.factor <-sqrt(nrow(dahms95.pca$x) -1)
d <- dahms95.pca$sdev
u <- sweep(dahms95.pca$x,2,1/(d*nobs.factor), FUN = "*")
v <- dahms95.pca$rotation

# scores
df.u <- as.data.frame(sweep(u,2,d^obs.scale, FUN = "*"))

# Directions
v <- sweep(v,2,d^var.scale, FUN = "*")
df.v <- as.data.frame(v)

df.u <- df.u * nobs.factor

# Scale the radius for the correlation circles so that it corresponds to the data ellipse for the standardized PC scores

r <- sqrt(qchisq(circle.prob, df = 2)) * prod(colMeans(df.u^2))^(1/4)

# Scale directions
v.scale <- rowSums(v^2)
df.v <- r * df.v/ sqrt(max(v.scale))
df.v %>% rownames_to_column("species") -> df.v

soi <- c("Poa pratensis", "Conyza canadensis", "Carex brevior", "Medicago lupulina", "Bromus japonicus")

ggplot(data = df.u) + 
  geom_point(aes(x = PC1, y = PC2)) + 
  geom_text(aes(x = PC1, y = PC2, label = paste(rownames(df.u))), alpha = 0.5, hjust = 1) +
  geom_segment(data = df.v %>% filter(species %in% soi), aes(x = 0, y = 0, xend = PC1, yend = PC2), arrow = arrow(length = unit(1/2, "picas")), color = muted("red")) +
  geom_label_repel(data = df.v %>% filter(species %in% soi), aes(x = PC1, y = PC2, label = species)) +
  theme_classic()

get_pca_var(dahms95.pca)
get_pca_ind(dahms95.pca)


```

## Nonmetric multidimensional scaling

NMDS has the best performance characteristics of the unconstrained ordination methods in ecology 

While most ordination methods attempt to maintain the distances between points in multivariate space, there are situations where the exact distance is immaterial and the rank order is sufficient. 

NMDS is able to use a multitude of distance measures

NMDS as opposed to other ordination methods does **NOT** rely on eigenanalysis and does not maximize the variation explained by each axis. 

NMDS axes are arbritrary and can be flipped and rotated

The number of NMDS axes are described *a priori*

The points are put into random space at the initial step.

The distances from the random locations are compared to the distances in the original data using isotonic regression with a stress function (stress ranges from 0 to 1)

The points are then iteratively moved in the direction of reduced stress until the lowest stress value is found 

```{r nmds}
dahms95.nmds <- metaMDS(dahms95, k = 2, try = 50) 
?metaMDS

dahms95.nmds
```

```{r plotnmds}
species_scores <- scores(dahms95.nmds, "species") %>% as.data.frame() %>% rownames_to_column("species")

year_scores <- scores(dahms95.nmds, "site") %>% as.data.frame() %>% rownames_to_column("year")


ggplot() + 
  # geom_point(data = species_scores, aes(x = NMDS1, y = NMDS2), shape = 16, colour = "red") +
  # geom_text(data = species_scores, aes(x = NMDS1, y = NMDS2, label = species), alpha = 0.25, hjust = 0) +
   geom_point(data = year_scores, aes(x = NMDS1, y = NMDS2), shape = 24, colour = "blue") + 
  theme_classic()
```

```{r fittingenviron}

precip_data %>% 
  filter(month(DATE) >= 4 & month(DATE) <=6,
         year(DATE) %in% c(2002:2017)) %>% 
  mutate(year = year(DATE)) %>% 
  group_by(year) %>% 
  summarise(ttl_precip = sum(PRCP, na.rm = TRUE)) -> total_precip


edge.sf <- ordisurf(dahms95.nmds ~ total_precip$ttl_precip, plot = TRUE)

extract.xyz <- function(obj){
  xy <- expand.grid(x = obj$grid$x, y = obj$grid$y)
  xyz <- cbind(xy, c(obj$grid$z))
  names(xyz) <- c("x","y","z")
  return(xyz)
  }

contour.vals<- extract.xyz(obj = edge.sf)

head(contour.vals)

ggplot(data = contour.vals) + 
  geom_point(data = species_scores, aes(x = NMDS1, y = NMDS2), shape = 16, colour = "red", alpha =0.25) +
   geom_point(data = year_scores, aes(x = NMDS1, y = NMDS2), shape = 24, colour = "blue", fill = "blue", size = 2, alpha = 0.5) + 
  geom_text(data = year_scores, aes(x = NMDS1, y = NMDS2, label = year), alpha = 0.45, hjust = 0) +
  geom_path(data = year_scores, aes(x = NMDS1, y = NMDS2), linetype = "dashed", arrow = arrow()) +
  geom_contour(aes(x = x, y = y, z = z), colour = "grey", alpha = 1) +
  coord_cartesian(xlim = c(-0.6,0.6), ylim = c(-0.4,0.4)) +
  theme_classic()

```

